---
title: 'Assignment #0 RMD'
author: "Michael Mayne"
date: "2026-01-25"
output: html_document
---

This project is for Assignment#0 as the starting deliverable for Data 607.

**Pre- Coding Approach**

For this assignment I am going to go over the basics to upload and load a data frame for reproducibility and general use. I plan on using a data set from NYC open data as they tend to be simpler to comprehend and data often comes with a clear data dictionary unlike a lot of other sources. The data set I plan to use is NYC Wifi Hot Spot Locations.

I will download the file on my local drive,then use filter, arrange functions to organize the data. Then i will drop an "unnecessary" columns such as the X, Y coordinates, Community Board numbers, NTA(Neighborhood Tabulation Area) and any non-descriptive identifiers. They are important but they do not display the information we need to understand the overview of the distribution of WiFi spots.

```{r}
# Loading Essential Libraries

library(dplyr)
library(tidyverse)

```

Introduction to Data

This data set is an overview of the available Wifi hotspots that individuals will be able to explore around the city.The data contains 3319 observations and 29 identifying categories. It was collect and placed on NYC open data. The last update as of this assignment was September 23 2002.

```{r}

#Upload Dataset from Github

Wifi_NYC <- read_csv("https://raw.githubusercontent.com/Mayneman000/DATA607Assignment/refs/heads/DATA/NYC_Wi-Fi_Hotspot_Locations_20260125.csv")

show_col_types = FALSE
```

```{r}

#Revealing the start of the Dataset
head(Wifi_NYC)
```

